Efficient Computation of Extended Sources

### 2021-05-14 repro/papers/sseprecond/paper.tex
interim

For acoustic wave physics, the coefficient vector is
$\bf{c}=(\kappa,\rho)$, with components bulk modulus $\kappa$ and
density $\rho$, and the state vector $\bu=(p,\bv)^T$ consists of
pressure $p$ (a scalar space-time field) and particle velocity $\bv$
(a vector space-time field). The wave operator $L[\bf{c}]$ is:
\begin{equation}
\label{eqn:aweop}
L[\bf{c}]\bf{u} = 
\left(
\begin{array}{c}
\frac{1}{\kappa}\frac{\partial p}{\partial t}  + \nabla \cdot \bv, \\
\rho\frac{\partial \bv}{\partial t} + \nabla p.
\end{array}
\right) 
\end{equation}
That is,
\begin{equation}
  \label{eqn:awemat}
  L[\bf{c}] = \left(
    \begin{array}{cc}
      \frac{1}{\kappa}\frac{\partial}{\partial t} & \nabla \cdot \\
      \nabla & \rho \frac{\partial}{\partial t}
    \end{array}
  \right)
\end{equation}

@@

Since all of the operators in the discussion that follows depend on
$\bf{c}$, I will suppress it from the notation, for example, $L=L[\bf{c}]$.

The extended ``crosswell'' configuration involves sources confined to
a spatial hypersurface. For convenience, choose this to be the hyperplane
$\{(x,y,z,t): z=z_s\}$. Such sources are combinations of constitutive law defects and
loads normal to the surface, that is, right-hand sides in the system
$L\bu=\bff$ of the form $\bff(\bx,t) =
(h_s(x,y,t)\delta(z-z_s), f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar
defect $h_s$ and normal force $f_s$  ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in \ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$ takes the form
\begin{eqnarray}
\label{eqn:awepm}
\frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} & = & - \nabla \cdot \bv^{\pm} +
h_s \delta(z-z_s), \nonumber \\
\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
\bv^{\pm} & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}

\noindent {\bf Remark:} In system \ref{eqn:awepm} and many similar
systems to follow, I will use the shorthand
\[
  p^+ = 0 \mbox{ for } t \ll 0
\]
to mean
\[
  \mbox{For some } T \in \bR, p^+(\cdot,t) = 0 \mbox{ for all } t \ge
  T.
\]
Extended forward modeling consists in solving \ref{eqn:awepm} and
sampling the solution components on another spatial hyperplane $\{(x,y,z,t): z=z_r\}$. Denote by $P_s,P_r$ the trace (sampling) operators on $z=z_s$, $z=z_r$
respectively. The causal/anti-causal modeling operators
${\cal S}^{\pm}_{z_s,z_r}$ are defined in
terms of the solutions $(p^{\pm},\bv^{\pm})$ of the systems
\ref{eqn:awepm} by
\begin{equation}
  {\cal S}^{\pm}_{z_s,z_r}(h_s,f_s)  = (P_rp^{\pm},P_r v_z^{\pm}),
  \label{eqn:fwd}
\end{equation}
The subscript signifies that sources are located on $z=z_s$, the
receivers on $z=z_r$. It is necessary to include this information in
the notation, as versions of ${\cal S}^{\pm}$ with sources and receivers in
several locations will be needed in the discussion below.

@@

Define $\Pi_i, i=0,1$ to be the projection on the first,
respectively second, component of a vector in $\bR^2$. Define the
forward modeling operator from pressure source to pressure trace by
\begin{equation}
  \label{eqn:sdef}
  S^{\pm}_{z_s,z_r} = \Pi_0 {\cal S}^{\pm}_{z_s,z_r} \Pi_0^T 
\end{equation}
and the forward modeling operator from velocity source (normal force)
to velocity trace by 
\begin{equation}
  \label{eqn:vdef}
  V^{\pm}_{z_s,z_r} = \Pi_1 {\cal S}^{\pm}_{z_s,z_r} \Pi_1^T 
\end{equation}
The relation \ref{eqn:trtr} implies that
\begin{eqnarray}
  (S^{\pm}_{z_s,z_r})^T &=& -S^{\mp}_{z_r,z_s} \nonumber\\
                        &=& R S^{\pm}_{z_r,z_s}R, \nonumber\\
    (V^{\pm}_{z_s,z_r})^T &=& -V^{\mp}_{z_r,z_s} \nonumber\\
                        &=& R V^{\pm}_{z_r,z_s}R.
                            \label{eqn:trtrcomp}
\end{eqnarray}

@@

Recall (equations \ref{eqn:deflam}, \ref{eqn:jump0}) that
\[
  \Lambda^+_{z_s} P_s p^+ = 2 \lim_{z \rightarrow z_s^+}v^+_z 
\]
and that with $h_s= \Lambda^+_{z_s}P_sp^+$,
\[
  S^+_{z_s,z_r}h_s = P_r p^+
\]
Together with the last two equations, the relation
\ref{eqn:approxinv0} implies that
\begin{equation}
  \label{eqn:approxinv1}
  S^+_{z_s,z_r}(2 P_s \tilde{v}_z^-) = P_rp^+.
\end{equation}
As shown in the last section, the solution of the system
\ref{eqn:revawe} differs negligibly from the anti-causal solution $(p^-_2,\bv^-_2)$
of the system \ref{eqn:sawe2} with (i) $z_s$ replaced by $z_r$, and
(ii) $\psi = [p^-_2]|_{z=z_r} = -2 lim_{z \rightarrow z_r^-} p^-_2
\approx -2 P_rp^+$. Therefore (definitions \ref{eqn:sdef}, \ref{eqn:vdef})
\begin{equation}
  \label{eqn:approxinv2}
  P_s \tilde{v}_z^- \approx V^-_{z_r,z_s}(-2 P_rp^+).
\end{equation}
Taken together, relations \ref{eqn:approxinv1} and
\ref{eqn:approxinv2} yield the first equation in the first main result of this section:
\begin{eqnarray}
  \label{eqn:approxinv}
  V^-_{z_r,z_s} S^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  S^-_{z_r,z_s} V^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  V^+_{z_s,z_r} S^-_{z_r,z_s} & \approx & \frac{1}{4}I, \nonumber\\
  S^+_{z_s,z_r} V^-_{z_r,z_s} & \approx & \frac{1}{4}I.\\
\end{eqnarray}.
The second equation follows by an exactly analogous argument, and the
last two by time reversal and interchange of the roles of $z_s$ and
$z_r$.

Having identified approximate two-sided inverses for $S^{\pm},
V^{\pm}$, the approximate null space identity \ref{eqn:lamnull} leads
to a relation between $\Lambda^{\pm}$ and ${\cal S}^{\pm}$ that
provides the foundation for practical computations. Applying $\Pi_0$
to both sides of \ref{eqn:lamnull} gives
\[
 \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_0^T
 \Lambda^+_{z_s} = S^+{z_s,z_r} \Lambda^+_{z_s}.
\]
From the first approximate identity in \ref{eqn:approxinv},
\begin{equation}
  \label{eqn:lamident}
  V^{-}_{z_r,z_s}\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \frac{1}{4}\Lambda^+_{z_s}.
\end{equation}
This identity is the second major result of this section: it shows how
to compute that action of $\Lambda^+_{z_s}$ by propagating the input
pressure trace, identified as a source for the velocity evolution,
forward in time from $z_s$ to $z_r$
reading off the pressure trace on $z=z_r$, identifying it once more as
a source for velocity, propagating it backwards in time from $z_r$ to
$z_s$, and finally reading off the velocity trace, interpreted as a
pressure evolution source on $z_s$. I will give a detailed description
of this computation in the section on numerical results.

@@

Computation of the transpose of $\Lambda^+$ (exact, not approximate in
the high frequency sense) will be important in the
construction of a preconditioner. The relation \ref{eqn:lamident} also
provides an computation for this operator. The idea is to treat the
left-hand side of equation \ref{eqn:lamident} as an approximation to
$\Lambda$, and compute its exact transposes via the adjoint state
method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
\begin{equation}
  \label{eqn:lamtransp}
   \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r} \approx \frac{1}{4}(\Lambda^+_{z_s})^T.
\end{equation} 
The left-hand sides of \ref{eqn:lamident} and \ref{eqn:lamtransp} will
simply be substituted for $\Lambda$ and its transpose in the
preconditioner construction: preconditioners must be (precisely)
self-adjoint, but need only be approximated to perform well.

>>>>

For acoustic wave physics, the coefficient vector is
$\bf{c}=(\kappa,\rho)^T$, with components bulk modulus $\kappa$ and
density $\rho$, and the state vector $\bu=(p,\bv)^T$ consists of
pressure $p$ (a scalar space-time field) and particle velocity $\bv$
(a vector space-time field). The wave operator $L[\bf{c}]$ is:
\begin{equation}
\label{eqn:aweop}
L[\bf{c}]\bf{u} = 
\left(
\begin{array}{c}
\frac{1}{\kappa}\frac{\partial p}{\partial t}  + \nabla \cdot \bv, \\
\rho\frac{\partial \bv}{\partial t} + \nabla p.
\end{array}
\right) 
\end{equation}
That is,
\begin{equation}
  \label{eqn:awemat}
  L[\bf{c}] = \left(
    \begin{array}{cc}
      \frac{1}{\kappa}\frac{\partial}{\partial t} & \nabla \cdot \\
      \nabla & \rho \frac{\partial}{\partial t}
    \end{array}
  \right)
\end{equation}

@@

Since all of the operators in the discussion that follows depend on
the coefficient vector 
$\bf{c}$, I will suppress it from the notation, for example, $L=L[\bf{c}]$.

The surface source extension replaces point sources on or near a
surface in $\bR^3$ with source functions confined to the same
surface. The simplest example of this extended geometry specifies a
plane $\{(x,y,z,t): z=z_s\}$ at source depth $z_s$ as the surface. For
acoustic modeling, surface sources are combinations of constitutive law
defects and loads normal to the surface, localized on $z=z_s$. That
is, right-hand sides in the system $L\bu=\bff$ take the form
$\bff(\bx,t) = (h_s(x,y,t)\delta(z-z_s),
f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar defect $h_s$ and normal
force $f_s$ ($\bf{e}_z=(0,0,1)$). With the choice $L$ given in
\ref{eqn:awemat}, the causal/anti-causal wave system $L\bu^{\pm}=\bff$
takes the form
\begin{eqnarray}
\label{eqn:awepm}
\frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} & = & - \nabla \cdot \bv^{\pm} +
h_s \delta(z-z_s), \nonumber \\
\rho\frac{\partial \bv^{\pm}}{\partial t} & = & - \nabla p^{\pm} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
p^{\pm} & =& 0 \mbox{ for } \pm t \ll 0,\nonumber\\ 
\bv^{\pm} & = & 0 \mbox{ for } \pm t \ll 0.
\end{eqnarray}

\noindent {\bf Remark:} In system \ref{eqn:awepm} and many similar
systems to follow, I will use the shorthand
\[
  p^+ = 0 \mbox{ for } t \ll 0 
\]
to mean that $p^+$ is {\em causal}, that is,
\[
  \mbox{For some } T \in \bR, p^+(\cdot,t) = 0 \mbox{ for all } t <
  T.
\]
Similarly,
\[
  p^- = 0 \mbox{ for } t \gg 0 
\]
signifies that $p^-$ is anti-causal.

Extended forward modeling consists in solving \ref{eqn:awepm} and
sampling the solution components at receiver locations. For
simplicity, throughout this paper I will assume that the receivers are
located on another spatial hyperplane $\{(x,y,z,t): z=z_r\}$ at
receiver depth $z_r>z_s$.

Denote by $P_s,P_r$ the trace (sampling)
operators on $z=z_s$, $z=z_r$ respectively.
Under the standing assumptions to be made on
the solutions of system \ref{eqn:awepm}, the key components of the
solution ($p^{\pm}$ and $v^{\pm}_z$) are continuous functions of $z$
in the open slab $z_s<z<z_r$, but may be discontinuous at the source
plane $z=z_s$. Similarly, the roles of $z_s$ and $z_r$ will be
interchanged in some of the constructions to come, and the
corresponding solutions may be discontinuous at $z=z_r$. Accordingly,
interpret $P_s$, $P_r$ as the limit from right and left respectively:
for $u=p^{\pm}$ or $v^{\pm}_z$,
\begin{eqnarray}
  \label{eqn:defsamp}
  P_su(x,y,t) &=& \lim_{z \rightarrow z_s^+} u(x,y,z,t),\nonumber \\
  P_ru(x,y,t) &=& \lim_{z \rightarrow z_r^-} u(x,y,z,t).                  
\end{eqnarray}

The causal/anti-causal vector
modeling operators ${\cal S}^{\pm}_{z_s,z_r}$ are defined in terms of
the solutions $(p^{\pm},\bv^{\pm})$ of the systems \ref{eqn:awepm} by
\begin{equation}
  {\cal S}^{\pm}_{z_s,z_r}(h_s,f_s)^T  = (P_rp^{\pm},P_r v_z^{\pm})^T,
  \label{eqn:fwd}
\end{equation}
The subscript signifies that sources are located on $z=z_s$, the
receivers on $z=z_r$. It is necessary to include this information in
the notation, as versions of ${\cal S}^{\pm}$ with sources and receivers in
several locations will be needed in the discussion below.

As will be shown later, ${\cal S}^{\pm}$ is not stably invertible, in fact
having rank 1 (in an approximate sense). The diagonal components of
${\cal S}^{\pm}$ thus carry essentially all of its information.
Denote by $\Pi_i, i=0,1$ the projection on the first,
respectively second, component of a vector in $\bR^2$. The 
forward modeling operator from pressure source to pressure trace is
\begin{equation}
  \label{eqn:sdef}
  S^{\pm}_{z_s,z_r} = \Pi_0 {\cal S}^{\pm}_{z_s,z_r} \Pi_0^T 
\end{equation}
and the forward modeling operator from velocity source (normal force)
to velocity trace is
\begin{equation}
  \label{eqn:vdef}
  V^{\pm}_{z_s,z_r} = \Pi_1 {\cal S}^{\pm}_{z_s,z_r} \Pi_1^T 
\end{equation}

@@

The relation \ref{eqn:trtr} implies that
\begin{eqnarray}
  (S^{\pm}_{z_s,z_r})^T &=& -S^{\mp}_{z_r,z_s} \nonumber\\
                        &=& R S^{\pm}_{z_r,z_s}R, \nonumber\\
    (V^{\pm}_{z_s,z_r})^T &=& -V^{\mp}_{z_r,z_s} \nonumber\\
                        &=& R V^{\pm}_{z_r,z_s}R.
                            \label{eqn:trtrcomp}
\end{eqnarray}

@@

Recall (equations \ref{eqn:deflam}, \ref{eqn:jump0}) that
\[
  \Lambda^+_{z_s} P_s p^+ \approx 2 \lim_{z \rightarrow z_s^+}v^+_z 
\]
and that with $h_s= \Lambda^+_{z_s}P_sp^+$,
\[
  S^+_{z_s,z_r}h_s = P_r p^+
\]
Together with the last two equations, the relation
\ref{eqn:approxinv0} implies that
\begin{equation}
  \label{eqn:approxinv1}
  S^+_{z_s,z_r}(2 P_s \tilde{v}_z^-) \approx P_rp^+.
\end{equation}
As shown in the last section, the solution of the system
\ref{eqn:revawe} differs negligibly from the anti-causal solution $(p^-_2,\bv^-_2)$
of the system \ref{eqn:sawe2} with (i) $z_s$ replaced by $z_r$, and
(ii) $\psi = [p^-_2]|_{z=z_r} = -2 \lim_{z \rightarrow z_r^-} p^-_2
\approx -2 P_rp^+$. Therefore (definitions \ref{eqn:sdef}, \ref{eqn:vdef})
\begin{equation}
  \label{eqn:approxinv2}
  P_s \tilde{v}_z^- \approx V^-_{z_r,z_s}(-2 P_rp^+).
\end{equation}
Taken together, relations \ref{eqn:approxinv1} and
\ref{eqn:approxinv2} yield the first equation in the first main result of this section:
\begin{eqnarray}
  \label{eqn:approxinv}
  V^-_{z_r,z_s} S^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
  S^-_{z_r,z_s} V^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
  V^+_{z_s,z_r} S^-_{z_r,z_s} & \approx & -\frac{1}{4}I, \nonumber\\
  S^+_{z_s,z_r} V^-_{z_r,z_s} & \approx & -\frac{1}{4}I.\\
\end{eqnarray}.
The second equation follows by an exactly analogous argument, and the
last two by time reversal and interchange of the roles of $z_s$ and
$z_r$.

Having identified approximate two-sided inverses for $S^{\pm},
V^{\pm}$, the approximate null space identity \ref{eqn:lamnull} leads
to a relation between $\Lambda^{\pm}$ and ${\cal S}^{\pm}$ that
provides the foundation for practical computations. Applying $\Pi_0$
to both sides of \ref{eqn:lamnull} gives
\[
 \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_0^T
 \Lambda^+_{z_s} = S^+_{z_s,z_r} \Lambda^+_{z_s}.
\]
From the first approximate identity in \ref{eqn:approxinv},
\begin{equation}
  \label{eqn:lamident}
  V^{-}_{z_r,z_s}\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx -\frac{1}{4}\Lambda^+_{z_s}.
\end{equation}
This identity is the second major result of this section: it shows how
to compute that action of $\Lambda^+_{z_s}$ by propagating the input
pressure trace, identified as a source for the velocity evolution,
forward in time from $z_s$ to $z_r$
reading off the pressure trace on $z=z_r$, identifying it once more as
a source for velocity, propagating it backwards in time from $z_r$ to
$z_s$, and finally reading off the velocity trace, interpreted as a
pressure evolution source on $z_s$. I will give a detailed description
of this computation in the section on numerical results.

@@

Computation of the transpose of $\Lambda^+$ (exact, not approximate in
the high frequency sense) will be important in the
construction of a preconditioner. The relation \ref{eqn:lamident} also
provides an computation for this operator. The idea is to treat the
left-hand side of equation \ref{eqn:lamident} as an approximation to
$\Lambda$, and compute its exact transposes via the adjoint state
method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
\begin{equation}
  \label{eqn:lamtransp}
   \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r} \approx -\frac{1}{4}(\Lambda^+_{z_s})^T.
\end{equation} 
The left-hand sides of \ref{eqn:lamident} and \ref{eqn:lamtransp} will
simply be substituted for $\Lambda$ and its transpose in the
preconditioner construction: preconditioners must be (precisely)
self-adjoint, but need only be approximated to perform well.

\section{Approximate Unitarity}
Recall that the fields $(p^{\pm},\bv^{\pm})$ in $z>z_s$ can be viewed
as either the solution of the system \ref{eqn:awepm} with pressure source $h_s=0$ and
a suitable choice of velocity source $f_s$, or of the same system with
suitable choice of $h_s$ and $f_s=0$, and that these choices are
related by equation \ref{eqn:lamfh}: $\Lambda^{\pm}f_s = h_s$. The
pressure traces at $z=z_r>z_s$ are the same: from the definitions
\ref{eqn:sdef}, \ref{eqn:vdef},
\begin{eqnarray}
  \label{eqn:sandv}
  S^{\pm}_{z_s,z_r}h_s& =& P_rp^{\pm},\nonumber \\
  V^{\pm}_{z_s,z_r}f_s &=& P_rv_z^{\pm}.
\end{eqnarray}
Applying the same reasoning to traces on $z=z_r$: create pressure and
velocity sources $h_r$ and $f_r$ at $z=z_r$ by odd extension of the
velocity, respectively pressure, fields, and using the jumps as source
coefficients. Then as before
$\Lambda^{\pm}_{z_r}f_r=h_r=-P_rv^{\pm}_z$ (the sign change is due to
$Lambda$ producing the jump, whereas $P_rv^{\pm}_z$ is the limit from
the left of the field $v^{\pm}_{1,z}$ of system \ref{eqn:awe1}), hence
$\Lambda^{\pm}_{z_r}P_rp^{\pm} = P_rv_z^{\pm}$. Combining this
observation with equations
\ref{eqn:lamfh}, \ref{eqn:sandv}, obtain (with roles of $z_s,z_r$ interchanged)
\[
V^{\pm}_{z_r,z_w}(-2P_rp^{\pm}) =
  \Lambda^{\pm}_{z_s}S^{\pm}_{z_r,z_s}(-2P_rv_z^{\pm})  =
  \Lambda^{\pm}_{z_s}S^{\pm}_{z_r,z_s}(-2\Lambda^{\pm}_{z_r}P_rp^{\pm})
\]
whence
\begin{equation}
  \label{eqn:preunit}
 V^{\pm}_{z_r,z_w}
 =\Lambda^{\pm}_{z_s}S^{\pm}_{z_r,z_s}\Lambda^{\pm}_{z_r}.
\end{equation}

In view of the relations \ref{eqn:trtrcomp}, equation
\ref{eqn:preunit} implies that
\[
  V^-_{z_r,z_s}=-(V^+_{z_s,z_r})^T = -\Lambda^{-}_{z_s}(S^+_{z_s,z_r})^T\Lambda^{-}_{z_r}.
\]
\[
  =-(\Lambda^{+}_{z_s})^T(S^+_{z_s,z_r})^T(\Lambda^{+}_{z_r})^T.
\]
Therefore the first equation in display \ref{eqn:approxinv} can be re-written
\[
  -4V^-_{z_r,z_s}S^+_{z_s,z_r} =
\]
\begin{equation}
  \label{eqn:baseunit}
 4 (\Lambda^{+}_{z_s})^T(S^+_{z_s,z_r})^T(\Lambda^{+}_{z_r})^T
  S^+_{z_s,z_r} \approx I
\end{equation}
